---
title: "How Economic Vulnerability Undermined Military Strategy in the 2025 U. S. Aid Cuts"
author: "Aryah, Nguyen, Jordyn"
date: "Date: `r Sys.Date()`"
format: 
  docx:
    #reference-doc: ../quarto_template.docx
    number-sections: true
    number-depth: 3
    fig-dpi: 500
    fig-align: center
    fig-height: 4
    fig-width: 6
execute: 
  echo: false
  warning: false
  message: false
---

## Introduction

How do factors such as inflation, democracy scores, and U.S. military involvement influence the depth of U.S. aid cuts to recipient countries? U.S. foreign aid has long played an important role in addressing poverty, food insecurity, and public health challenges worldwide. When aid declines sharply, aided governments may struggle to provide essential services like medicine, clean water, and agricultural support. Research on recent foreign aid cuts shows that reductions often have immediate downstream effects on health systems, food security, and economic stability in receiving countries (Better World Campaign). Because significant cuts have occurred, understanding why some countries face larger reductions than others is important. It can reveal the real priorities shaping U.S. foreign policy and international partnerships, which can help us understand and improve the program in the future.

In this project, we examine three factors that may help explain why some recipient countries experience deeper U.S. aid cuts than others. To motivate our selection of factors, we reference "Who Gives Foreign Aid to Whom and Why?" researched by Alesina and Dollar. In their study, they determine that aid disbursement is primarily driven by strategic and political factors, not just need (Alesina & Dollar, 2000). This suggests that aid cuts, much like allocations, may reflect a balance between humanitarian concerns and broader political priorities. First, we consider inflation as a measure of economic vulnerability. Inflation rates may matter because countries with high inflation usually have less economic stability and a greater need for outside help. Second, as Alesina and Dollar explore, political alignment may shape aid decisions, since governments that differ strongly from U.S. political values might be viewed as less important partners. If this is the case, countries that align less closely with the U.S. democratic system may be more vulnerable to aid cuts. Finally, we include U.S. military presence to capture a country's strategic importance. Countries hosting a large number of U.S. troops might hold greater strategic importance and, therefore, be protected from sharp budget reductions. Together, these factors represent a country’s economic need, political alignment, and security importance to the United States, allowing us to understand whether U.S. aid cuts in 2025 followed a pattern or reflected trade-offs between humantiarian and strategic priorities.

## Data and Design

The data for this project come from several separate sources that provide information on U.S. aid, inflation rates, democracy scores, population sizes, and military presence. Specifically, the U.S. aid disbursement data comes from [*foreignassistance.gov*](foreignassistance.gov), while macroeconomic indicators like population and inflation were sourced from [*globalmacrodata.com*](globalmacrodata.com). Our political and military variables were accessed via R packages, with democracy scores coming from the Varieties of Democracy (V-Dem) Polyarchy Index via [`{isard}`](https://svmiller.com/isard/index.html) and troop deployments via [`{troopdata}`](https://meflynn.github.io/troopdata/). A strength of combining these sources is that it provides a comprehensive, multi-dimensional view of recipient countries. However, a weakness is that merging different databases often results in missing data for certain countries, which reduces the number of observations available for our statistical tests and may limit how broadly our findings apply.

The dependent variable measures the depth of aid cuts, defined as the change in U.S. aid disbursements between 2024 and 2025, so that larger negative values indicate larger reductions. The explanatory variables capture three factors that could have influenced this cut. Inflation is measured using 2024 inflation transformed with the inverse hyperbolic sine, which helps compare countries with very different inflation levels while preserving meaningful interpretation for small values. Democracy assessment is measured through a country’s vdem score. Finally, troop density calculated as the number of U.S. troops stationed in a country divided by its 2024 population, which will also be transformed with the inverse hyperbolic sine.

These predictor values are transformed with `asinh` to reduce skew and improve comparability across countries. We will first graph the distribution of variables to justify these transformations and ensure they are appropriate. We will then use a coefficient plot with confidence intervals to show the magnitude and uncertainty of each predictor. Finally, we will show prediction plots in which the change in aid is modeled as a function of military presence, inflation, and political polarization. All statistical inference relies on robust standard errors to account for potential differences in error variability across countries.

## Analysis

```{r}
## open packages:
library(tidyverse) # for data wrangling tools and data viz
library(estimatr)  # for model estimation with robust standard errors
library(sjPlot)    # for making prediction plots
library(cowplot)   # for combining plots

## the aid dataset:
dt <- read_csv("https://raw.githubusercontent.com/milesdwilliams15/Teaching/refs/heads/main/DPR%20201/Data/dpr201_unit1_data.csv")

## Data wrangling & Feature Engineering
analysis_data <- dt %>%
  mutate(
    # Dependent Variable: Change in Asinh Disbursements
    aid_cut = asinh(disbursed_2025) - asinh(disbursed_2024),
    
    # Factor 1: Military Density
    military_density = us_troops / pop_2024,
    asinh_military_density = asinh(military_density),
    
    # Factor 2: Inflation
    asinh_inf_2025 = asinh(inf_2025),
    
    # Factor 3: Political Polarization
    gov_polarization = abs(vdem_score - 0.5)) %>%
  drop_na(aid_cut, asinh_military_density, asinh_inf_2025, gov_polarization) %>%
  filter(is.finite(aid_cut))

## Set plot theme globally
set_theme(base = theme_minimal())

## Estimate the model
model <- lm_robust(aid_cut ~ asinh_military_density + asinh_inf_2025 + gov_polarization,                    data = analysis_data, 
                   se_type = "HC1")
model_smry <- tidy(model)
```

Before evaluating our regression model, we must visually inspect our continuous variables to ensure our transformations were effective. Figures 1 and 2 display the density plots for military presence and inflation, respectively. In their raw forms, both variables are highly right-skewed; the vast majority of countries cluster near zero, while a handful of extreme outliers stretch the x-axis. This skew would violate the assumptions of a linear model. By applying the inverse hyperbolic sine (`asinh`) transformation, we successfully compress these long tails and normalize the distributions, making the data appropriate for regression analysis.

```{r}
# Figure 1: Military density transformation
p1 <- ggplot(analysis_data, aes(x = military_density)) +
  geom_density(fill = "lightblue", alpha = 0.4) +
  labs(title = "Figure 1: Distribution of Military Density", subtitle =  "Raw",x = "Troops per Million", y = "Density") +
  theme_minimal()

p2 <- ggplot(analysis_data, aes(x = asinh_military_density)) +
  geom_density(fill = "lightblue", alpha = 0.4) +
  labs(title = "", subtitle = "Transformed", x = "Asinh(Troops per Million)", y = "") +
  theme_minimal()

plot_grid(p1, p2, nrow = 1)
```

```{r}
# Figure 2: Inflation 2025 transformation
p3 <- ggplot(analysis_data, aes(x = inf_2025)) +
  geom_density(fill = "lightcoral", alpha = 0.4) +
  labs(title = "Figure 2: Distribution of Inflation 2025",subtitle =  "Raw", x = "Inflation (%)", y = "Density") +
  theme_minimal()

p4 <- ggplot(analysis_data, aes(x = asinh_inf_2025)) +
  geom_density(fill = "lightcoral", alpha = 0.4) +
  labs(title = "", subtitle =  "Transformed", x = "Asinh(Inflation)", y = "") +
  theme_minimal()

plot_grid(p3, p4, nrow = 1)

```

Now that the data is balanced, we run our linear regression model to test our three factors. Figure 3 shows the estimated effect of each factor, along with lines indicating our level of certainty (confidence intervals) and statistical significance (p-values). The result for military density is negative and statistically significant (p = 0.006). This indicates that countries hosting a higher density of U.S. troops unexpectedly faced deeper cuts to their aid. Conversely, the result for inflation is positive and significant (p = 0.012), demonstrating that the U.S. protected nations experiencing severe economic instability from the worst cuts. Political alignment (polarization) is not statistically significant (p = 0.386), meaning the political ideology of the recipient government did not systematically influence the budget decisions.

```{r}
# Figure 3: Coefficient Plot
model_smry %>%
  filter(term != "(Intercept)") %>%
  ggplot(aes(
    x = estimate,
    xmin = conf.low,
    xmax = conf.high,
    y = term
  )) +
  geom_pointrange(color = "darkblue") +
  # Add the p-values above the points
  geom_text(aes(label = paste0("p = ", round(p.value, 3))), 
            vjust = -1.5, 
            size = 3.5, 
            color = "black") +
  geom_vline(xintercept = 0, linetype = 2) +
  labs(
    title = "Figure 3: Predictors of 2025 Aid Cuts",
    x = "Coefficient Estimate (with 95% Confidence Intervals)",
    y = NULL
  ) +
  scale_y_discrete(labels = c(
    "gov_polarization" = "Government Polarization",
    "asinh_inf_2025" = "Inflation (Asinh)", 
    "asinh_military_density" = "Military Density (Asinh)"
  ))
```

To visualize the substantive magnitude of these findings, Figure 4 provides marginal prediction plots for each variable. Panel B illustrates a *clear penalty* for military hosts: as the density of U.S. troops increases, the expected change in aid drops sharply into negative numbers (deeper cuts). Panel C highlights a *positive upward slope*, indicating that higher inflation correlates with protected (less negative) aid payments. Finally, Panel A remains *relatively flat*, visually confirming that the U.S. did not base its budget decisions on whether a nation was a strong democracy or an autocracy.

```{r}
# Figure 4: Prediction Plots
# Generate individual plots

# Figure 4: Prediction Plots

# Generate individual plots with short sub-titles
p1 <- plot_model(model, type = "pred", terms = "gov_polarization", 
                 title = "A", 
                 axis.title = c("Government Polarization", "Expected Aid Cuts"))

p2 <- plot_model(model, type = "pred", terms = "asinh_military_density", 
                 title = "B", 
                 axis.title = c("Troops per million (Asinh)", ""))

p3 <- plot_model(model, type = "pred", terms = "asinh_inf_2025", 
                 title = "C", 
                 axis.title = c("Inflation 2025 (Asinh)", ""))

# Combine the three plots into one row
plot_row <- plot_grid(p1, p2, p3, nrow = 1)

# Create a standalone title object
main_title <- ggdraw() + 
  draw_label("Figure 4: Predicting U.S. Aid Cuts", x = 0.3, vjust = 1)

# Combine the title and the row of plots
plot_grid(main_title, plot_row, ncol = 1, rel_heights = c(0.1, 1))
```

## Conclusion

This analysis sought to determine whether a specific logic drove the depth of U.S. foreign aid cuts in 2025. By testing the impact of military presence, economic instability, and political alignment, our findings show that the reductions were neither random nor based on political values. Instead, the results suggest that economic conditions played the most consistent role in determining aid cuts in 2025. Countries experiencing higher inflation saw the lowest amount of aid cuts, while our political affiliation variable of governmental polarization did not meaningfully shape aid cuts. Countries with larger U.S. troop presence experienced deeper reductions in aid, contradicting our previous hypothesis that U.S. military presence may reflect strategic importance, and therefore, be shielded from aid cuts.

Unexpectedly, nations heavily reliant on the U.S. military (those with high troop density) experienced the sharpest cuts in aid. This pattern suggests the U.S. administration may have viewed these nations as capable of absorbing the financial hit, or as the politically safest targets for immediate budget savings. Research on public support for foreign assistance indicates that aid is especially vulnerable to cuts when its effectiveness is questioned or when its purpose is unclear to general audiences, even in strategically important countries (Hurst, Tidwell, & Hawkins, 2017). On the other hand, the U.S. actively spared countries suffering from severe inflation, acting as a financial safety net for the most economically vulnerable partners. Moreover, policymakers should be cautious about interpreting aid reductions in countries with U.S. military presence as a direct reflection of weakening strategic importance. Instead, these findings may suggest that short-term pressures and the current administration's political goals may drive aid cuts more than long-term strategic importance. Further, the significance of protecting high-inflation countries implies that future policy decisions may prioritize economic stabilization, reflecting a more humanitarian viewpoint. Ultimately, these results suggest that in 2025, American foreign aid operated primarily as an economic first-responder rather than a tool to reward military or political allies.

\newpage

## References

Alesina, A., & Dollar, D. (2000). Who Gives Foreign Aid to Whom and Why? Journal of Economic Growth, 5(1), 33–63. http://www.jstor.org/stable/40216022

Better World Campaign. (n.d.). The impact of foreign assistance cuts. Better World Campaign. https://betterworldcampaign.org/impact-of-foreign-assistance-cu ts

Hurst, R., Tidwell, T., & Hawkins, D. (2017). Down the Rathole? Public Support for US Foreign Aid. International Studies Quarterly, 61(2), 442–454. http://www.jstor.org/stable/44510179

\newpage

## Code Appendix

<!-- Don't touch this! This lets me see your code at the end of the rendered document. -->

```{r ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE}
```
